sfislands: An R Package for Accommodating Islands and Disjoint Zones in Areal Spatial Modelling

### 2024-03-28 article.Rmd
reading models from github

liverpool_hmodel <- readRDS(here("data","liverpool_model2.rds"))

@@

liverpool_hsamodel <- readRDS(here("data","liverpool_hsamodel.rds"))

@@

liverpool_hsamodel_dist_band <- readRDS(here("data","liverpool_hsamodel_dist_band.rds"))

>>>>

# read locally
# liverpool_hmodel <- readRDS(here("data","liverpool_model2.rds"))

# read from github
liverpool_hmodel <- readRDS(gzcon(url("https://github.com/horankev/r_journal_sfislands/raw/main/data/liverpool_model2.rds")))

@@

# read locally
# liverpool_hsamodel <- readRDS(here("data","liverpool_hsamodel.rds"))

# read from github
liverpool_hsamodel <- readRDS(gzcon(url("https://github.com/horankev/r_journal_sfislands/raw/main/data/liverpool_hsamodel.rds")))

@@

# read locally
# liverpool_hsamodel_dist_band <- readRDS(here("data","liverpool_hsamodel_dist_band.rds"))

# read from github
liverpool_hsamodel_dist_band <- readRDS(gzcon(url("https://github.com/horankev/r_journal_sfislands/raw/main/data/liverpool_hsamodel_dist_band.rds")))

### 2024-03-28 article.tex
reading models from github

\abstract{%
Fitting areal models which use a spatial weights matrix to represent relationships between geographical units can be a cumbersome task, particularly when these units are not well-behaved. The two chief aims of \texttt{sfislands} are to simplify the process of creating an appropriate neighbourhood matrix, and to quickly visualise the predictions of subsequent models. The package uses visual aids in the form of easily-generated maps to help this process. This paper demonstrates how \texttt{sfislands} could be useful to researchers. It begins by describing the package's functions in the context of a proposed workflow. It then presents three worked examples showing a selection of potential use-cases. These range from earthquakes in Indonesia, to river crossings in London, and hierarchical models of output areas in Liverpool. We aim to show how the \texttt{sfislands} package streamlines much of the human workflow involved in creating and examining such models.
}

@@

A key feature which differentiates spatial statistics is the
non-independence of observations and the expectation that neighbouring
units will be more similar than non-neighbouring ones (Tobler 1970). If this
is not accounted for, the assumptions of many types of models will be
violated. The relationships between all spatial units in a study can be
represented numerically in a spatial weights matrix. In order to build
this, we must first decide on what constitutes being a neighbour. We can
see this as a continuous relationship where degree of neighbourliness is
a function of connectivity, which could be represented as some measure
of distance. Alternatively it could be a binary situation where each
pair of units either are (1) or are not (0) neighbours. This can be
based on a condition such as contiguity of some sort, or a distance
constraint. It is the job of the modeller to formulate a hypothesis
which justifies their choice of neighbourhood structure.

For R users, the \CRANpkg{spdep} package (Bivand et al. 2005) has long been popular for the
creation of these matrices. More recently, with the increasing use of \CRANpkg{sf} (Pebesma 2018)
structures and their compatibility with the \CRANpkg{tidyverse} (Wickham et al. 2019), the \CRANpkg{sfdep} package (Parry and Locke 2024) has presented the same
functionality in addition to extra features in a tidy structure based on
lists in a similar way to \CRANpkg{sf}.

@@

The most commonly used neighbourhood structure is one based on first-order queen contiguity, where units are considered neighbours if they
share at least a vortex of boundary. However, as the name suggests, this
will lead to problems when non-contiguous units such as islands or
exclaves are present. Less obviously, depending on how the geographic
units are described, areas on either sides of rivers may be
inappropriately classified as neighbours or not neighbours. Furthermore,
the presence of infrastructure such as tunnels, bridges or ferry
services might be satisfactory to meet our hypothesis of the required
degree of connectivity to be considered neighbours. Again, such information will may not be apparent from a basic set of polygons. In order to create what a researcher considers to be an appropriate neighbourhood structure, incorporating all of the domain knowledge that they might have about the system, it should be simple and intuitive to add and remove connections between spatial units. This might mean adding links to account for man-made infrastructure, or cutting links to incorporate natural barriers such as rivers or mountains.

The aim of \texttt{sfislands} is to deal with the situations described above in a
convenient and open manner. It allows us to set up a structure, quickly
map it, and then examine whether or not we are happy with how it represents our hypothesis of relationships between units. The structure can then be edited and the process re-iterated until we have described a
spatial relationship structure with which we are satisfied.

@@

In this paper, we will look at three examples to show different use-cases for
\texttt{sfislands}. The first example focuses on earthquakes in
Indonesia. It shows a scenario where all of the functions are used, from
setting up contiguities, to modelling and examining the predictions of the
model.

The second example looks at London and how, despite the lack of islands,
the presence of a river means that some of the pre-functions of
\texttt{sfislands} can be useful.

The final example focuses on Liverpool. There are no islands or issues
of discontiguity in this dataset. Instead, this example shows how the package can be used to quickly fit a set of reasonably complex multilevel models, with or without a spatially autocorrelated component at the lowest level, and how the results can be presented in tidy format and quickly visualised. This example also gives an opportunity to show how, when different types of contiguity from, say, the \CRANpkg{spdep} package are used in an otherwise equivalent model, the \texttt{sfislands} post-functions make it very straightforward to quickly compare the different predictions generated from different structures.

\hypertarget{why-use-sfislands}{%
\section{\texorpdfstring{Why use \texttt{sfislands}?}{Why use sfislands?}}\label{why-use-sfislands}}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  It addresses an issue commonly seen in online help forums where an inexperienced user wishes to get started with a model but fails at the first hurdle because their neighbourhood structure is not compatible with their data. \texttt{sfislands} will include a contiguity for all units.
\item
  It gives tools to immediately visualise this structure as a map.
\item
  These maps are created using \CRANpkg{ggplot2} (Wickham 2016), which allows users to apply the desired styling and themes using \CRANpkg{ggplot2} syntax.
\item
  As the nodes can be labelled by index, it makes it very easy to add and remove connections as appropriate with confidence.
\item
  Connections which have been induced by a function from the package but are not contiguities can be accessed to ensure openness in the process.
\end{enumerate}

@@

These structures can serve as the input to models in \CRANpkg{brms},
\pkg{r-inla}, \pkg{rstan} (Stan Development Team 2020) or \CRANpkg{mgcv}. \CRANpkg{brms} requires a matrix structure while \CRANpkg{mgcv} models use a list. Rather than having a separate neighbours
object, it is included in the original \CRANpkg{sf} dataframe as a named list or matrix, in the spirit of the \CRANpkg{sfdep} package.

@@

Having decided upon an appropriate neighbourhood structure, the next step is to use this in the context of a model. The use of such structures is particularly associated with CAR-type (conditional autoregressive) models.
These are often implemented in a Bayesian framework using
\CRANpkg{brms}, \pkg{r-inla} or \pkg{rstan}. For example, the \CRANpkg{brms} ICAR structure requires the neighbourhood relationships to be in matrix form. The pre-functions will output the
neighbourhood structure in the desired format for use in any of these
frameworks. A convenient frequentist alternative is to use the
\CRANpkg{mgcv} package which requires a named list of neighbours. It has the functionality to create such models
using \texttt{bs="mrf"}. It also has the ability to combine these with a
hierarchical structure using \texttt{bs="re"}. While the outputs from the
Bayesian structures mentioned above can be extracted in the same way as
any other component of the model, it can be somewhat awkward to get the
estimates from \CRANpkg{mgcv} model. \texttt{sfislands} has two post-functions to conveniently extract and visualise these.

@@

This function augments the original dataframe with the estimated means
and standard errors of the spatially varying predictions from a fitted \CRANpkg{mgcv} model in a similar manner to
how the \CRANpkg{broom} package (Robinson, Hayes, and Couch 2023) operates. The \texttt{geometry} column, as per convention, remains as the last column of the augmented dataframe, while the predictions are positioned immediately before it. \footnote{In a similar way, \texttt{st\_augment()} can also be used to append the random effects from \CRANpkg{lme4} (Bates et al. 2015) and \CRANpkg{nlme} (Pinheiro, Bates, and R Core Team 2023) models to an \texttt{sf} dataframe, which can then be easily mapped using \texttt{st\_quickmap\_preds()}. Compatability with models created using different packages can be introduced in the future.} The spatially varying predictions which \texttt{st\_augment()} extracts from an \CRANpkg{mgcv} model are

@@

Consider the model structure described in the code below using \CRANpkg{mgcv} syntax. In this model \(y\) is the dependent variable while it is being estimated with a fixed intercept, a fixed slope for some covariate, a set of random intercepts and slopes for the covariate at a \emph{region} level, and a set of ICAR varying intercepts and slopes at a lower \emph{sub-region} level.

@@

\hypertarget{example-1-indonesia}{%
\section{Example 1: Indonesia}\label{example-1-indonesia}}

@@

Having described the type of model we wish to implement, we now show how \texttt{sfislands} can be used to streamline the process.

@@

In this case, we use \texttt{st\_bridges()} for setting up the queen contiguity
structure as usual, but with the additional stipulation that unconnected
units (provinces which are islands or collections of islands) are
considered neighbours to their \emph{k} nearest provinces. For this example,
we have set the value of k to 2. The resulting neighbourhood structure
is shown in Figure \ref{fig:ind-contig1}.

@@

Regardless of this, having controlled for the concentration of faults, what is the additional risk level of earthquakes in a province? This can be seen as a measure of the activity level of faults locally and it is spatially smoothed by the autoregressive process. It is represented in the model summary by the component \texttt{s(province)}. However, the extraction of individual predictions for this component for each province from the \texttt{mgcv} model requires a number of steps. We now demonstrate how these are streamlined into a single function by the \texttt{sfislands} package.

@@

The function \texttt{st\_augment()} allow us to add the spatially varying predictions from the model as new columns to the original dataframe in a process
similar to that of the \CRANpkg{broom} package. For instance, we see from the code below that the original dataframe is now augmented with columns called ``mrf.smooth.province'' and ``se.mrf.smooth.province'' which show the predictions for the \(\gamma_i\) component and their standard errors. Note that this is how we would expect them to be named, based on the previous discussion surrounding Table \ref{tab:staugtab-latex}. They are positioned immediately before the final \texttt{geometry} column of the \texttt{sf} dataframe.

@@

In this example, we have gone through a number of stages carefully, making changes to contiguities that we deemed appropriate as we went. However, in practice, at least in a first iteration, it might not be necessary to go through all of these steps. A rough and ready model, complete with spatially varying coefficients and visual output, can be generated with \texttt{sfislands} using nothing more than three or four lines of code, such as the following:

@@

\hypertarget{example-2-london}{%
\section{Example 2: London}\label{example-2-london}}

The next example looks only at using the \emph{pre-functions} of \texttt{sfislands}, but in a situation where the presence of actual \emph{islands} is not the problem we seek to deal with. Consider the wards and boroughs of London (sourced from the Greater London Authority's \href{https://data.london.gov.uk/}{London Datastore}). In Figure \ref{fig:lon-contig1} the \texttt{st\_bridges()}
function is applied to them to construct a queen contiguity
neighbourhood structure. Because there are no disjoint units (or
``islands''), this will be the same as using \texttt{st\_contiguity()} from
\CRANpkg{sfdep}. The \texttt{st\_quickmap\_nb()} function gives an immediate
visual representation of the structure. \footnote{\texttt{st\_quickmap\_nb()} can also be used to visualise any contiguity structure created by \texttt{spdep} or \texttt{sfdep} as long as that structure is included in an \texttt{sf} dataframe as a column named \texttt{nb}.} Because this map is created using \CRANpkg{ggplot2}, it can be easily supplemented by adding a layer showing the course of the river Thames which is also visible in Figure \ref{fig:lon-contig1}.

@@

When a study area has a river running through it, problems can arise with constructing appropriate neighbourhood structures. Depending on how the
geometries are defined, the presence of a river can cause problems in two ways. In one situation, the river could be expressed as a polygon in its own right meaning that it severs any potential contiguity between units on either side of its banks. In this situation, no spatial units will be neighbours with the units directly across the river from them. At the other extreme, if the river is not included as a geometry (as is the case here) all units on opposing banks are considered neighbours.

Depending on the presence of river crossings, two areas which are physically quite close but on opposing banks might be very distinct. If there is no means of crossing the river within a reasonable distance, somebody living on the banks of a river might be more likely to go about their life primarily on their side of the river, despite the short distance as the crow flies of facilities on the other side. This could be relevant in terms of, say, modelling of house prices where we might want to incorporate these issues into a neighbourhood structure.

\texttt{sfislands} provides convenient functions for this sort of situation. Let us start by restricting our wards and boroughs of interest to just those which are on either side of the river Thames (Figure \ref{fig:lon-contig2}).

@@

This allows us to easily cut the ties across the river for these units
manually by using the function \texttt{st\_manual\_cut\_nb()}. \footnote{While we are using the index of the units in this example, the function also
  accepts names as arguments which may be more convenient in some circumstances.} Having made these adjustments,
\texttt{st\_quickmap\_nb()} now shows a connectivity structure (Figure \ref{fig:lon-contig4}) which reflects how we have hypothesised that influence should extend across the river in the presence or absence of crossings.

This example shows that the pre-functions of \texttt{sfislands} have uses for situations which do not involve islands. They can be used to apply domain knowledge to easily design the most appropriate neighbourhood structure.

@@

\hypertarget{example-3-liverpool}{%
\section{Example 3: Liverpool}\label{example-3-liverpool}}

In this final example, there are no islands and no river issues. We saw in the Indonesia example how \texttt{sfislands} can extract estimates from the \texttt{mgcv} argument \texttt{bs="mrf"}. It can also deal with random effects as defined by \texttt{bs="re"} in \texttt{mgcv} models. This means it is possible to create quite complex models featuring both a traditional multi-level structure and conditionally related components. Here, we will construct a hierarchical model in this way which also features an ICAR component at the lowest level of the tree. We will compare two different types of contiguity which could be applied to the ICAR component:

\begin{itemize}
\tightlist
\item
  \texttt{st\_bridges()} from \texttt{sfislands}, which implements first-order queen contiguity in the same way as \texttt{sfdep:st\_contiguity()} for situations such as this where there are no islands, and
\item
  \texttt{st\_dist\_band()} from \CRANpkg{sfdep}, which connects units whose centroids lie within a chosen distance threshold.
\end{itemize}

@@

There are no islands or discontiguities in this geometry of Liverpool so using \texttt{st\_bridges()} will give the same structure as would \texttt{sfdep::st\_contiguity()} (see Figure \ref{fig:liv-contigs}).

@@

To demonstrate how \texttt{sfislands} can be used to quickly compare the output of models, we show the output of three similar models. The first of these is a hierarchical model as described above, but with no ICAR component at the lowest level. Its spatially varying coefficients are shown in Figure \ref{fig:liv-est1}. Secondly, we fit the model described above using \texttt{st\_bridges()} to generate first-order queen contiguities as the structure for the neighbourhood relationships. The coefficients of this model are shown in Figure \ref{fig:liv-est2}.

@@

This shows how \texttt{sfislands} can quickly give an intuitive representation of how a set of similar models are operating. As a measure of the relative strengths of these three structures, we could compare their AICs (see Table \ref{tab:aic-latex}). These suggest that, in this scenario, inclusion of the ICAR component is appropriate, and queen contiguity is a more suitable choice than the distance-band condition for defining the neighbourhood structure.

@@

\caption{\label{tab:aic-latex}Comparison of models by AIC.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{l|r|r}
\hline
\textbf{model} & \textbf{AIC} & \textbf{relative AIC}\\
\hline
hierachical model, ICAR based on st\_bridges() & -5126.948 & -1.000\\
\hline
hierachical model, ICAR based on sfdep::st\_dist\_band() & -5107.790 & -0.996\\
\hline
hierachical model, no ICAR & -5082.684 & -0.991\\
\hline
\end{tabular}
\end{table}

@@

These examples have shown the varying scenarios in which \texttt{sfislands} can be useful. It aims to contribute to spatial modelling by making an awkward area less awkward. Rather than having a default attitude of ignoring islands when building neighbourhood structures based on contiguity, it hopes to encourage at least an examination of whether or not it is appropriate for them to be included. It also provides helper functions to use these structures in spatial regression models built with \CRANpkg{mgcv} which streamline the human effort necessary to examine the estimates. In future, compatibility with other modelling packages can be added to broaden the package's capabilities.

@@

\leavevmode\vadjust pre{\hypertarget{ref-spdep}{}}%
Bivand, Roger, Andrew Bernat, Marilia Carvalho, Yongwan Chun, Carsten Dormann, Stéphane Dray, Rein Halbersma, et al. 2005. {``The Spdep Package.''} \emph{Comprehensive R Archive Network, Version}, 05--83.

@@

\leavevmode\vadjust pre{\hypertarget{ref-ggpubr}{}}%
Kassambara, Alboukadel. 2023. \emph{Ggpubr: 'Ggplot2' Based Publication Ready Plots}. \url{https://rpkgs.datanovia.com/ggpubr/}.

\leavevmode\vadjust pre{\hypertarget{ref-sfdep}{}}%
Parry, Josiah, and Dexter H Locke. 2024. \emph{Sfdep: Spatial Dependence for Simple Features}. \url{https://sfdep.josiahparry.com}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-wikicrossings}{}}%
Wikipedia. 2024. {``{List of crossings of the River Thames} --- {W}ikipedia{,} the Free Encyclopedia.''} \url{http://en.wikipedia.org/w/index.php?title=List\%20of\%20crossings\%20of\%20the\%20River\%20Thames\&oldid=1184426738}.

>>>>

\abstract{%
Fitting areal models which use a spatial weights matrix to represent relationships between geographical units can be a cumbersome task, particularly when these units are not well-behaved. The two chief aims of \CRANpkg{sfislands} are to simplify the process of creating an appropriate neighbourhood matrix, and to quickly visualise the predictions of subsequent models. The package uses visual aids in the form of easily-generated maps to help this process. This paper demonstrates how \CRANpkg{sfislands} could be useful to researchers. It begins by describing the package's functions in the context of a proposed workflow. It then presents three worked examples showing a selection of potential use-cases. These range from earthquakes in Indonesia, to river crossings in London, and hierarchical models of output areas in Liverpool. We aim to show how the \CRANpkg{sfislands} package streamlines much of the human workflow involved in creating and examining such models.
}

@@

A key feature which differentiates spatial statistics is the
non-independence of observations and the expectation that neighbouring
units will be more similar than non-neighbouring ones (Tobler 1970). If this
is not accounted for, the assumptions of many types of models will be
violated. The relationships between all spatial units in a study can be
represented numerically in a spatial weights matrix. In order to build
this, we must first decide on what constitutes being a neighbour. We might
see this as a continuous relationship where degree of neighbourliness is
a function of connectivity, which could be represented as some measure
of distance. Alternatively it could be a binary situation where each
pair of units either are (1) or are not (0) neighbours. This can be
based on a condition such as contiguity of some sort, or a distance
constraint. It is the job of the modeller to formulate a hypothesis
which justifies their choice of neighbourhood structure.

For R users, the \CRANpkg{spdep} package (Bivand, Hauke, and Kossowski 2013) has long been popular for the
creation of these matrices. More recently, with the increasing use of \CRANpkg{sf} structures (Pebesma 2018) and their compatibility with the \CRANpkg{tidyverse} (Wickham et al. 2019), the \CRANpkg{sfdep} package (Parry 2023) has presented the same
functionality in addition to extra features in a tidy structure based on
lists in a similar way to \CRANpkg{sf}.

@@

The most commonly used neighbourhood structure is one based on first-order queen contiguity, where units are considered neighbours if they
share at least a vortex of boundary. However, as the name suggests, this
will lead to problems when non-contiguous units such as islands or
exclaves are present. Less obviously, depending on how the geographic
units are described, areas on either sides of rivers may be
inappropriately classified as neighbours or not neighbours. Furthermore,
the presence of infrastructure such as tunnels, bridges or ferry
services might be satisfactory to meet our hypothesis of the required
degree of connectivity to be considered neighbours. Again, such information may not be apparent from a basic set of polygons. In order to create what a researcher considers to be an appropriate neighbourhood structure, incorporating all of the domain knowledge that they might have about the system, it should be simple and intuitive to add and remove connections between spatial units. This might mean adding links to account for man-made infrastructure, or cutting links to incorporate natural barriers such as rivers or mountains.

The aim of \CRANpkg{sfislands} (Horan, Domijan, and Brunsdon 2024) is to deal with the situations described above in a
convenient and open manner. It allows us to set up a structure, quickly
map it, and then examine whether or not we are happy with how it represents our hypothesis of relationships between units. The structure can then be edited and the process re-iterated until we have described a
spatial relationship structure with which we are satisfied.

@@

In this paper, we will look at three examples to show different use-cases for
\CRANpkg{sfislands}. The first example focuses on earthquakes in
Indonesia. It shows a scenario where all of the functions are used, from
setting up contiguities, to modelling and examining the predictions of the
model.

The second example looks at London and how, despite an absence of islands,
the presence of a river means that some of the pre-functions of
\CRANpkg{sfislands} can be useful.

The final example focuses on Liverpool. There are no islands or issues
of discontiguity in this dataset. Instead, this example shows how the package can be used to quickly fit a set of reasonably complex multilevel models, with or without a spatially autocorrelated component at the lowest level, and how the results can be presented in tidy format and quickly visualised. This example also gives an opportunity to show how, when different types of contiguity from, say, the \CRANpkg{spdep} package are used in an otherwise equivalent model, the \CRANpkg{sfislands} post-functions make it very straightforward to quickly compare the different predictions generated from different structures.

\hypertarget{why-use}{%
\section{\texorpdfstring{Why use \CRANpkg{sfislands}?}{Why use ?}}\label{why-use}}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  It addresses an issue commonly seen in online help forums where an inexperienced user wishes to get started with a model but fails at the first hurdle because their neighbourhood structure contains empty records. \CRANpkg{sfislands} will include a contiguity for all units.
\item
  It gives tools to immediately visualise this structure as a map.
\item
  These maps are created using \CRANpkg{ggplot2} (Wickham 2016), which allows users to apply additional styling and themes using \CRANpkg{ggplot2} syntax.
\item
  As the nodes can be labelled by index, it makes it very easy to add and remove connections as appropriate with confidence.
\item
  Connections which have been induced by a function from the package but which are not based on geographical contiguity can be accessed to ensure openness in the process.
\end{enumerate}

@@

These structures can serve as the input to models in \CRANpkg{brms},
\pkg{r-inla}, \CRANpkg{rstan} (Stan Development Team 2020) or \CRANpkg{mgcv}. \CRANpkg{brms} requires a matrix structure while \CRANpkg{mgcv} models use a list. Rather than having a separate neighbours
object, it is included in the original \CRANpkg{sf} dataframe as a named list or matrix, in the spirit of the \CRANpkg{sfdep} package.

@@

Having decided upon an appropriate neighbourhood structure, the next step is to use this in the context of a model. The use of such structures is particularly associated with CAR (conditional autoregressive) or ICAR-type (intrinsic conditional autoregressive) models (Besag 1974).
These are often implemented in a Bayesian framework using
\CRANpkg{brms}, \pkg{r-inla} or \CRANpkg{rstan}. For example, the \CRANpkg{brms} ICAR structure requires the neighbourhood relationships to be in matrix form. The pre-functions will output the
neighbourhood structure in the desired format for use in any of these
frameworks. A convenient frequentist alternative is to use the
\CRANpkg{mgcv} package which requires a named list of neighbours. It has the functionality to create such models
using \texttt{bs="mrf"}. It also has the ability to combine these with a
hierarchical structure using \texttt{bs="re"}. While the outputs from the
Bayesian structures mentioned above can be extracted in the same way as
any other component of the model, it can be somewhat awkward to get the
estimates from \CRANpkg{mgcv} models. \CRANpkg{sfislands} has two post-functions to conveniently extract and visualise these.

@@

This function augments the original dataframe with the estimated means
and standard errors of the spatially varying predictions from a fitted \CRANpkg{mgcv} model in a similar manner to
how the \CRANpkg{broom} package (Robinson, Hayes, and Couch 2023) operates. The \texttt{geometry} column, as per convention, remains as the last column of the augmented dataframe, while the predictions are positioned immediately before it. \footnote{In a similar way, \texttt{st\_augment()} can also be used to append the random effects from \CRANpkg{lme4} (Bates et al. 2015) and \CRANpkg{nlme} (Pinheiro, Bates, and R Core Team 2023) models to an \CRANpkg{sf} dataframe, which can then be easily mapped using \texttt{st\_quickmap\_preds()}. Compatability with models created using different packages can be introduced in the future.} The spatially varying predictions which \texttt{st\_augment()} extracts from an \CRANpkg{mgcv} model are

@@

Consider the model structure described in the code below using \CRANpkg{mgcv} syntax. In this model \(y\) is the dependent variable which is being estimated with a fixed intercept, a fixed slope for some covariate, a set of random intercepts and slopes for the covariate at a \emph{region} level, and a set of ICAR varying intercepts and slopes at a lower \emph{sub-region} level.

@@

\hypertarget{indonesia-example-1}{%
\section{Indonesia (example 1)}\label{indonesia-example-1}}

@@

Having described the type of model we wish to implement, we now show how \CRANpkg{sfislands} can be used to streamline the process.

@@

In this case, we use \texttt{st\_bridges()} for setting up the queen contiguity
structure as usual, but with the additional stipulation that unconnected
units (provinces which are islands or collections of islands) are
considered neighbours to their \emph{k} nearest provinces. For this example,
we have set the value of k to 2. The resulting neighbourhood structure
is shown in Figure \ref{fig:ind-contig1}. Note how is can be styled with a combination of internal arguments (size, colour, fill etc.) and additional \CRANpkg{ggplot2} layers.

@@

Returning to the initial question, what is the additional risk level of earthquakes in a province, having controlled for the concentration of faults? This can be seen as a measure of the activity level of faults locally and it is spatially smoothed by the autoregressive process. It is represented in the model summary by the component \texttt{s(province)}. However, the extraction of individual predictions for this component for each province from the \texttt{mgcv} model requires a number of steps. We now demonstrate how these are streamlined into a single function by the \CRANpkg{sfislands} package.

@@

The function \texttt{st\_augment()} allow us to add the spatially varying predictions from the model as new columns to the original dataframe in a process
similar to that of the \CRANpkg{broom} package. For instance, we see from the output of the following code chunk that the original dataframe is now augmented with columns called \texttt{mrf.smooth.province} and \texttt{se.mrf.smooth.province} which show the predictions for the \(\gamma_i\) component and their standard errors. Note that this is how we would expect them to be named, based on the previous discussion surrounding Table \ref{tab:staugtab-latex}. They are positioned immediately before the final \texttt{geometry} column of the \texttt{sf} dataframe, and after the neighbours list column, \texttt{nb}.

@@

In this example, we have gone through a number of stages carefully, making changes to contiguities that we deemed appropriate as we went. However, in practice, at least in a first iteration, it might not be necessary to go through all of these steps. A rough and ready model, complete with spatially varying coefficients and visual output, can be generated with \CRANpkg{sfislands} using nothing more than three or four lines of code, such as the following:

@@

\hypertarget{london-example-2}{%
\section{London (example 2)}\label{london-example-2}}

The next example looks only at using the \emph{pre-functions} of \CRANpkg{sfislands}, but in a situation where the presence of actual \emph{islands} is not the problem we seek to deal with. Consider the wards and boroughs of London (sourced from the Greater London Authority's \href{https://data.london.gov.uk/}{London Datastore}) and available at
\url{https://github.com/horankev/london_liverpool_data}. In Figure \ref{fig:lon-contig1} the \texttt{st\_bridges()}
function is applied to them to construct a queen contiguity
neighbourhood structure. Because there are no disjoint units (or
``islands''), this will be the same as using \texttt{st\_contiguity()} from
\CRANpkg{sfdep}. The \texttt{st\_quickmap\_nb()} function gives an immediate
visual representation of the structure. \footnote{\texttt{st\_quickmap\_nb()} can also be used to visualise any contiguity structure created by \texttt{spdep} or \texttt{sfdep} as long as that structure is included in an \texttt{sf} dataframe as a column named \texttt{nb}.} Because this map is created using \CRANpkg{ggplot2}, it can be easily supplemented by adding a layer showing the course of the river Thames which is also visible in Figure \ref{fig:lon-contig1}.

@@

When a study area has a river running through it, problems can arise with constructing appropriate neighbourhood structures. Depending on how the
geometries are defined, the presence of a river can cause problems in two ways. In one situation, the river could be expressed as a polygon in its own right meaning that, using the condition of queen contiguity, it severs any potential contiguity between units on either side of its banks. In this situation, no spatial units will be neighbours with the units directly across the river from them. At the other extreme, if the river is not included as a geometry (as is the case here) all units on opposing banks are automatically considered neighbours.

Depending on the presence of river crossings, two areas which are physically quite close but on opposing banks might be very distinct. If there is no means of crossing the river within a reasonable distance, somebody living on the banks of a river might be more likely to go about their life primarily on their side of the river, despite the short distance as the crow flies of facilities on the other side. This could be relevant in terms of, say, modelling of house prices where we might want to incorporate issues such as local amenities into a neighbourhood structure.

\CRANpkg{sfislands} provides convenient functions for this sort of situation. Let us start by restricting our wards and boroughs of interest to just those which are on either side of the river Thames. Figure \ref{fig:lon-contig2} shows the resultant contiguities when the river is ignored.

@@

This allows us to easily cut the ties across the river for these units
manually by using the function \texttt{st\_manual\_cut\_nb()}. \footnote{While we are using the index of the units in this example, the function also
  accepts names as arguments which may be more convenient in some circumstances.} Having made these adjustments,
\texttt{st\_quickmap\_nb()} now shows a connectivity structure (Figure \ref{fig:lon-contig4}) which reflects our hypothesis of how influence should extend across the river in the presence or absence of crossings.

This example shows that the pre-functions of \CRANpkg{sfislands} have uses for situations which do not involve islands. They can be used to apply domain knowledge to easily design the most appropriate neighbourhood structure.

@@

\hypertarget{liverpool-example-3}{%
\section{Liverpool (example 3)}\label{liverpool-example-3}}

In this final example, there are no islands and no river issues. We saw in the Indonesia example how \CRANpkg{sfislands} can extract estimates from the \texttt{mgcv} argument \texttt{bs="mrf"}. It can also deal with random effects as defined by \texttt{bs="re"} in \texttt{mgcv} models. This means it is possible to create quite complex models featuring both a traditional multi-level structure and conditionally correlated components. Here, we will construct a hierarchical model in this way which also features an ICAR component at the lowest level of the tree. We will compare two different types of contiguity which could be applied to the ICAR component:

\begin{itemize}
\tightlist
\item
  \texttt{st\_bridges()} from \CRANpkg{sfislands}, which implements first-order queen contiguity in the same way as \texttt{sfdep:st\_contiguity()} for situations such as this where there are no islands, and
\item
  \texttt{st\_dist\_band()} from \CRANpkg{sfdep}, which connects units whose centroids lie within a chosen distance threshold.
\end{itemize}

@@

There are no islands or discontiguities in this geometry of Liverpool so using \texttt{st\_bridges()} will give the same structure as would \texttt{sfdep::st\_contiguity()} (see Figure \ref{fig:liv-contigs}, with a magnified section for visibility purposes, created using \pkg{ggmagnify} (Hugh-Jones 2024)).

@@

To demonstrate how \CRANpkg{sfislands} can be used to quickly compare the output of models, we show the output of three similar models. The first of these is a hierarchical model as described above, but with no ICAR component at the lowest level. Its spatially varying coefficients are shown in Figure \ref{fig:liv-est1}. Secondly, we fit the model described above using \texttt{st\_bridges()} to generate first-order queen contiguities as the structure for the neighbourhood relationships. The coefficients of this model are shown in Figure \ref{fig:liv-est2}.

@@

This shows how \CRANpkg{sfislands} can quickly give an intuitive representation of how a set of similar models are operating. As a measure of the relative strengths of these three structures, we could compare their AICs (see Table \ref{tab:aic-latex}). These suggest that, in this scenario, inclusion of the ICAR component is appropriate, and queen contiguity is a more suitable choice than the distance-band condition for defining the neighbourhood structure.

@@

\caption{\label{tab:aic-latex}Comparison of models by AIC.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{l|r|r}
\hline
\textbf{model} & \textbf{AIC} & \textbf{relative AIC}\\
\hline
hierarchical model, ICAR based on st\_bridges() & -5126.948 & -1.000\\
\hline
hierarchical model, ICAR based on sfdep::st\_dist\_band() & -5107.790 & -0.996\\
\hline
hierarchical model, no ICAR & -5082.684 & -0.991\\
\hline
\end{tabular}
\end{table}

@@

These examples have shown the varying scenarios in which \CRANpkg{sfislands} can be useful. It aims to contribute to spatial modelling by making an awkward area less awkward. Rather than having a default attitude of ignoring islands when building neighbourhood structures based on contiguity, it hopes to encourage at least an examination of whether or not it is appropriate for them to be included. Even when no islands are present, it provides a simple procedure for tailoring a neighbourhood structure to match a given hypothesis. It also provides helper functions to use these structures in spatial regression models, notably those built with \CRANpkg{mgcv}, which streamline the human effort necessary to examine the estimates. In future, compatibility with other modelling packages can be added to broaden the package's capabilities.

@@

\leavevmode\vadjust pre{\hypertarget{ref-besag}{}}%
Besag, Julian. 1974. {``Spatial Interaction and the Statistical Analysis of Lattice Systems.''} \emph{Journal of the Royal Statistical Society: Series B (Methodological)} 36 (2): 192--225. \url{https://doi.org/10.1111/j.2517-6161.1974.tb00999.x}.

\leavevmode\vadjust pre{\hypertarget{ref-spdep}{}}%
Bivand, Roger, Jan Hauke, and Tomasz Kossowski. 2013. {``Computing the Jacobian in Gaussian Spatial Autoregressive Models: An Illustrated Comparison of Available Methods.''} \emph{Geographical Analysis} 45 (2): 150--79. \url{http://www.jstatsoft.org/v63/i18/}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-sfislands}{}}%
Horan, Kevin, Katarina Domijan, and Chris Brunsdon. 2024. \emph{{sfislands}: Streamlines the Process of Fitting Areal Spatial Models}. \url{https://horankev.github.io/sfislands/}.

\leavevmode\vadjust pre{\hypertarget{ref-ggmagnify}{}}%
Hugh-Jones, David. 2024. \emph{Ggmagnify: Create a Magnified Inset of Part of a "Ggplot" Object}. \url{https://github.com/hughjonesd/ggmagnify}.

\leavevmode\vadjust pre{\hypertarget{ref-ggpubr}{}}%
Kassambara, Alboukadel. 2023. \emph{{ggpubr}: {``{ggplot2}''} Based Publication Ready Plots}. \url{https://CRAN.R-project.org/package=ggpubr}.

\leavevmode\vadjust pre{\hypertarget{ref-sfdep}{}}%
Parry, Josiah. 2023. \emph{{sfdep}: Spatial Dependence for Simple Features}. \url{https://CRAN.R-project.org/package=sfdep}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-wikicrossings}{}}%
Wikipedia. 2024. {``{List of crossings of the River Thames} --- {W}ikipedia{,} the Free Encyclopedia.''} \url{http://en.wikipedia.org/w/index.php?title=List/\%20of/\%20crossings/\%20of/\%20the/\%20River/\%20Thames\&oldid=1184426738}.
